\chapter{Introduction}
\label{chap:intro}

\begin{chapabstract}

\textrm{{\bf Abstract:}} The present thesis has been driven typically by the development and investigation of machine learning methods in genomic data analysis for improved breast cancer survival prediction and for detecting robust genomic signatures related to cancer prognosis. Towards these goals, two directions in terms of methodologies have been explored for supervised survival prediction and structured feature selection, based on rank-based and network-guided approaches respectively. This chapter serves as a general introduction to the thesis that briefly summarizes the background of breast cancer prognosis and reviews the thesis work with a particular focus on how it addresses the computational challenges in genomic data analysis for cancer prognosis.
\linebreak
\vskip 0.1in
\noindent \textrm{{\bf Résumé:}}

\end{chapabstract}

\section{General Background}

Breast cancer refers to a malignant tumor that has developed from cells in the breast. Uncontrolled growth of cancer cells can invade nearby healthy breast tissue over time, and if cancer cells get into the lymph nodes that are small organs that filter out foreign substances in the body, they could then have a system of spreading further into other parts of the body and form new tumors in distant organs or tissues, a process called distant metastasis that aggravates the situation to a significant extent. Breast cancer is the most common cancer in women worldwide and second most common cancer overall for both genders in terms of incidence rates (after lung cancer). It is one of the leading causes of women's death from cancer, and over 508,000 women worldwide were estimated to have died in 2011 due to breast cancer (Global Health Estimates, WHO 2013). Survival rates have in general been improving over the past decade as a result of increased awareness, earlier detection through screening, adequate medical care and cancer treatment advances, with the caveat that rates vary greatly worldwide and still remain quite low in less developed countries.


If diagnosed early, the initial treatment for breast cancer is usually accomplished by complete removal of tumor by surgery or radiation (without damage to the rest of the body). After the initial treatment (or in case that the initial treatment should not be applicable), many patients are advised to receive systemic treatment including adjuvant chemotherapy, hormone therapy and targeted therapy to lower the risk of relapse, the recurrence risk of cancer-related conditions, and to prevent metastasis. The decision of receiving systemic treatment is made based on prognosis of the cancer patient, the estimation of the risk of relapse or likely course of outcome if no systemic treatment is given after the initial treatment.\footnote{In fact, two questions need to be addressed in decision making for cancer treatment: prognosis that is the identification of those patients who are most likely to benefit from the systemic treatment, and prediction that is the determination on which specific treatment should be most responsive and effective for the patients. While prognosis and prediction are equally important and usually discussed together in literature, prediction will be omitted from discussion for ease of the presentation of the thesis.} For example, as the most common type of systemic treatment, adjuvant chemotherapy usually involves cytotoxic drugs and has strong deleterious side effects. Consequently, such aggressive treatments are most beneficial for patients with poor prognosis but the intake of adjuvant chemotherapy should be minimized for those that do not necessarily benefit much from it. Therefore, accurate prognosis for each individual patient is of chief importance in order to personalize the best therapeutic option.


Conventionally, breast cancer prognosis is based solely on clinico-pathological information collected of patients and tumors. Several commonly used clinical-pathological parameters have been well-established to be indicative of likely prognosis of patients, and thus widely adopted in the clinical management of breast cancer. For example, it is known that breast cancer with cancer cells detected in lymph nodes has a higher risk of relapse than breast cancer \textit{in situ}, and thus requires to be treated with more aggressive adjuvant chemotherapies \cite{Moffat2014Clinical}. In fact, doctors most often evaluate the severity of breast cancer based on the Nottingham grading system, a score-based grading system using clinico-pathological parameters such as the size and shape of the nucleus in the tumor cells and how many dividing cells are present \cite{Cancer2010AJCC}. High-grade tumors look the most abnormal from normal cells and tend to be the most invasive, and are thus classified with poor prognosis. As another example, hormone receptors in breast cancer, estrogen-receptor (ER) and progesterone-receptor (PR), play an important role in normal glandular development and in breast cancer progression, and their status is therefore highly prognostic (as well as predictive to the responsiveness of hormone and endocrine therapies) \cite{Moffat2014Clinical}. Some online tools exist to perform prognosis of cancer patients and aid physicians weigh against the risks and benefits of adjuvant treatments, among which stands out the renowned \textit{Adjuvant! Online}\footnote{\url{https://www.adjuvantonline.com/}.}. Notably, the 6 predictors that are shown highly prognostic and used by \textit{Adjuvant! Online} to predict cancer-related mortality and relapse are: patient age, tumor size, grade, hormone receptor status, number of positive lymph nodes and comorbidity level.


Due to the intrinsic heterogeneity across breast cancer tumors, patients of similar clinico-pathological type can have remarkably different survival outcome. An example constituted by \cite{Veer2008Enabling} will be quoted here. Large meta-analyses show that recurrence is likely in 20--30\% of young women with early-stage (lymph node-negative) breast cancer, but in the United States 85--90\% of women with this type of cancer receive adjuvant chemotherapy, among whom 55--75\% therefore undergo a toxic therapy that they would very likely not benefit from but may experience the undesirable side effects. Since cancer is a inherently complex disease, the unwanted situation is mostly due to the fact that clinico-pathological information alone is far from sufficient to reliably identify those patients who are likely to relapse, let alone to accurately characterize the outcome of each particular case. It is recognized as an important yet challenging task to improve prognosis for each diseased individual and identify more efficient prognostic features, burgeoning the research of interest in interrogating breast cancer at the molecular level.




\section{Towards Molecular-level Breast Cancer Prognosis}

Among many theories on cancer biology, a widely accepted theory states that cancer is caused by genomic abnormalities, such as the accumulation of mutations or the dysregulation of gene expression involving tumor suppressor genes and oncogenes in cancer cells. It has long been established that genomic features contain unique characteristics of each individual being and offer the opportunity of scrutinizing the individuality of each breast tumor. Often termed by \textit{biomarkers} are such molecular-level features, typically genes, whose abnormal presence or dysfunctional behavior characterizes the biological heterogeneity of tumours, leading to molecular subtyping of cancer, and can thus be indicative of prognosis. While biomarkers can be associated to any phenotype of interest in general, the discussion will be restricted to biomarkers related to breast cancer prognosis in accordance with the objective of the present thesis.


Many biomarkers related to breast cancer survival have been reported in literature. For example, somatic mutations in gene TP53 show association with worse survival, independent of other risk predictors, see for instance a meta-analysis by \cite{Pharoah1999Somatic}. Worse breast cancer survival of gene BRCA mutation carriers versus non-carriers have been confirmed by several meta-analyses \cite{Zhong2015Effects, Zhu2016BRCA}. Over-expression of gene HER2, pathologically termed as HER2-positive, is linked to poorer outcome of node-negative breast cancers \cite{Chia2008Human}, a widely-observed association that has led to the advent of several HER2-directed therapies \cite{Arteaga2012Treatment}. Notably, major molecular subtypes of breast cancer are determined by the gene expression status, over- or under-expression, of hormone receptors and HER2, based on which physicians usually perform prognosis and plan treatments \cite{Schnitt2010Classification}. For a review on currently established and emerging biomarkers for breast cancer prognosis, see \cite{Weigel2010Current}.


The foundation of Human Genome Project (HGP) and the rapid advancement of genomic profiling technologies in the past decades have paved way to the advent of the current ``-omics'' revolution. Nowadays, thousands up to millions of genomic features can be efficiently collected from biological samples available for medical research. For example, in gene expression profiling, DNA microarray, a hybridization-based technology, measures the relative expression activity of a predetermined list of target genes \cite{Lockhart1996Expression}, and RNA-seq, a next-generation sequencing-based technology, was later invented to provide information on the expression of tens of thousands of gene sequences at lower cost and higher throughput (or larger genome coverage) with many advantages benchmarked against previous technologies \cite{Wang2009RNA}.


The revolution of gene expression profiling technologies fostered the development of multigene expression signatures for breast cancer prognosis, a group of biomarker genes whose combined expression pattern refines prognosis (usually with incremental value added to the use of classic clinico-pathological parameters alone). The research of prognostic signatures has resulted in many success stories \cite{Sotiriou2009Gene}. Notably, there exist at least 6 different prognostic multigene expression signatures commercially available to aid clinical decision making of breast cancer:\footnote{\url{http://www.breastcancer.org/symptoms/testing/types}.}
\begin{itemize}
	\item MammaPrint\textsuperscript{\textregistered} (Agendia, Amsterdam, The Netherlands) \cite{Veer2002Gene} is a 70-gene microarray-based expression profile for stratifying breast cancer into high- or low-risk prognostic groups. As one of the earliest success stories, it was the first test approved by the Food and Drug Administration (FDA) in the United States and by regulators in the European Union as an adjunct prognostic assay for breast cancer patients younger than 61 years of age with stage I/II, lymph node-negative or one to three lymph node-positive disease. The prognostic risk discrimination is good among ER-positive cancers but has limited value for ER-negative cancers.
	\item Prosigna\textsuperscript{\textregistered} Breast Cancer Prognostic Gene Signature Assay or PAM50 (Nanostring Technologies, Seattle, WA, USA) \cite{Parker2009Supervised} is a 50-gene qPCR assay for classifying breast tumors into five intrinsic subtypes (luminal A, luminal B, HER2-enriched, basal-like, normal-like) that are prognostic independent of standard clinico-pathological parameters. It is the second FDA-approved test in the United States to estimate distant recurrence risk for stage I/II (including one to three positive nodes), ER-positive breast cancer in postmenopausal women treated with adjuvant endocrine therapy, and it also received clearance in the European Union.
	\item Oncotype DX\textsuperscript{\textregistered} (Genomic Health, Redwood City, CA, USA) \cite{Paik2004multigene} is a 21-gene signature for categorizing adjuvant-tamoxifen-treated breast cancer patients into groups of low-, intermediate- or high-risk recurrence. It is the most widely used prognostic assay for ER-positive cancers in the United States.
	\item MapQuant Dx\textsuperscript{TM} Genomic Grade Index (Ipsogen, France) \cite{Sotiriou2006Gene} is a microarray-based 97-gene assay for reclassifying histologically intermediate-grade ER-positive cancers into high or low molecular grade with significantly different prognosis.
	\item Breast Cancer Index\textsuperscript{SM} (BioTheranostics, San Diego, CA, USA) \cite{Ma2008five} is comprised of two signatures, a five-gene molecular grade index and the ratio of two independent biomarkers HOXB13:IL17BR, and can predict the risk of distant recurrence in ER-positive, lymph node-negative breast cancers.
	\item EndoPredict\textsuperscript{\textregistered} (made by Sividon Diagnostics GmbH, Koln, Germany and distributed by Myriad Genetics, Salt Lake City, UT, USA) \cite{Filipits2011new} is a 11-gene signature for stratifying patients with ER-positive cancer into a low or a high risk of recurrence if treated with adjuvant endocrine therapy alone.
\end{itemize}
More details about these signatures are found in \cite{Gyorffy2015Multigene}.


In order to study the substantial amount of genomic data available for cancer research, the use of computational tools such as machine learning has become a popular trend \cite{Barillot2012Computational}. In fact, machine learning is particularly suitable for analyzing genomic data by developing algorithms or building models to discover unseen patterns, identify complex relationships and predict for phenotypic phenomenon of interest. For example, PAM50 test, the 50-gene classifier for breast cancer subtypes, is constructed upon a learning algorithm called the nearest shrunken centroid method \cite{Tibshirani2002Diagnosis}. Another successful application comes from \textit{DREAM 7 --- Sage Bionetworks--DREAM Breast Cancer Prognosis Challenge} \cite{Margolin2013Systematic}, a competition-based crowd-source effort that systematically assessed the potential of computational models designed to predict breast cancer survival and the contribution of various types of clinico-pathological and molecular features to improving prognosis performance \cite{Bilal2013Improving}. Notably, the best-performing model of the competition \cite{Cheng2013Development} was built upon, in addition to clinico-pathological features, such molecular features called ``attractor metagenes'' that are pan-cancer signatures of coexpressed genes previously identified in rich gene expression data sets by an iterative attractor-finding algorithm \cite{Cheng2013Biomolecular}. For a recent survey on machine learning applications in cancer prognosis, see \cite{Kourou2015Machine}.




\section{Genomic Data Analysis: Prospects and Challenges}

While genomic data analysis of cancer is a research field encompassing a broad range of topics, the present thesis is specifically devoted to breast cancer prognosis and related biomarker discovery. Towards that end, two major questions need to be addressed from the viewpoint of machine learning:

\paragraph{Supervised Learning for Survival Risk Prediction}

prognosis in the language of machine learning typically survival analysis and binary classification... survival analysis is blah on four types of survival : DMFS, RFS, DFS, OS (saini's thesis table 1.1); binary labels converted from survival are used in classification of higher- or lower- risk of survival, good vs poor prognosis... pure ML algorithms such as Cox model for former, SVM, RF, Boosting for latter...

\paragraph{Feature Selection for Biomarker Discovery}

whole genome vs feature selection... pure ML algorithm such as univariate filter, RFE, regularization...

challenges exist... it has been widely recognized as a very challenging problem to extract potentially valuable information from genomic data... Besides the intrinsic complexity of cancer biology and heterogeneity of patient data, relatively small number of clinical samples, the inherent measurement noise in high-throughput experimental data makes it demanding... particularly in biomarker discovery, robustness of the identified biomarkers and \textit{a posteriori} interpretation of the findings are also major concerns... to address special issues of genomic data, bio-oriented methods beyond but inspired by pure ML algorithm...

top scoring pairs...

cancer is a network disease... notably from genome to genes to networks... genome-wide study to biomarker discovery to network-level analysis...




\section{Contribution of the Thesis}

The present thesis is conceived following two lines of ideas intended to address the two major questions in genomic data analysis for cancer prognosis, and will be outlined by projects, each presented in one chapter.

\paragraph{Rank-based Approaches for Improved Survival Risk Prediction.}

The first line of ideas is to perform gene expression data analysis based solely on the ranking of the expression levels of multiple genes while their real-valued measurements are disregarded. It is based on the assumption that the relative ordering of gene expression levels can potentially be more robust and informative predictors than their real-valued measurements in biomedical applications since high-throughput high-dimensional genomic data are often subject to high measurement noise. From the point view of machine learning, it reduces to the study of a particular type of structured data, specifically rankings. It is well-known that kernel methods have found many successful applications where the input data are discrete or structured including strings and graphs. The first project of my doctoral studies was focused on proposing computationally attractive kernels for rank data and applying kernel methods to problems involving rankings. Central to this work was the observation that the widely used Kendall tau correlation and the Mallows similarity measures are indeed computationally attractive positive definite kernels for rankings. These kernels were further tailored for more complex types of rank data that prevail in real-world applications, especially when rankings come from real-valued vectors by keeping simply the relative ordering of the values of multiple features thereof. Thanks to these kernels, many off-the-shelf kernel machines are available to solve various problems at hand. It is worth special mention that, despite that the project was initially motivated by biomedical applications, the prospective contribution of this work concerns applications from many fields of machine learning pertaining to learning from rankings, or learning to rank. This study will be presented in Chapter \ref{chap:kendall}.


The study of the Kendall kernel for rankings has paved an unprecedented way towards a deeper understanding of a classical problem called Kemeny aggregation from the field of Social Choice Theory. Kemeny aggregation searches for a consensus ranking that best represents a collection of individual rankings in the sense that the sum of the Kendall tau distance between each ranking and the consensus is minimized. Kemeny aggregation is often considered to provide the ideal solution among all ranking aggregation criteria but the Kemeny consensus is known to be NP-hard to find. Many tractable approximations to the Kemeny consensus have therefore been proposed and well studied. Since the Kendall kernel derives from an inner product of a Euclidean space, the Kendall tau distance derives from a squared Euclidean distance and thus the combinatorial problem of Kemeny aggregation is endowed with a novel intuitive interpretation from a geometric point of view. Based on this observation, a tractable upper bound of the estimation error in terms of the distance between the Kemeny consensus and its approximation is established. This upper bound requires little assumption on the approximation procedure or the collection of rankings to aggregate. However, due to its remote connection to cancer prognosis that is the primary objective of the present thesis, this study will be presented in Appendix \ref{chap:kemeny}. 


\paragraph{Network-guided Approaches for Enhanced Biomarker Discovery.}

The second line of ideas of performing genomic data analysis for cancer prognosis is to consult biological networks as domain-specific knowledge to guide our analysis. In fact, biological networks are a common way of depicting functional relationships between genes that have been accumulated from decades of biomedical research, and they can be particularly valuable when incorporated as prior knowledge in order to improve the selection efficacy of biomarkers related to breast cancer prognosis. Typical examples of biological networks include protein-protein interaction network and signaling pathway network. Two projects were initiated on network-guided gene expression data analysis for breast cancer prognosis and for the identification of genomic signatures related to cancer survival. 


In one project, we focused on the study of structured regularization in generalized linear models for selecting prognostic genes where the regularization method is designed so that genes closer on the biological network are encouraged to be selected simultaneously. In fact, in order to achieve such simultaneous modularity and sparsity coherent with respect to presumed network structure, a popular method called network-based wavelet smoothing has been successfully applied in many applications from the field of signal processing. Therefore, we were motivated to investigate the potential of this method in biomedical applications, using a breast cancer gene expression dataset and a protein-protein interaction network to guide our analysis albeit the methodology is generally applicable to various types of genomic data and biological networks. In particular, the method allows to designate a few prognostic gene modules with intra-collaborative functionality rendering readily interpretable insights potentially related to cancer survival. Numerical results demonstrated that, compared to several network-free and some established network-based regularization methods, network-based wavelet smoothing was able to improve the selection efficacy of prognostic genes in terms of stability, connectivity and interpretability, while achieving competitive performance of survival risk prediction for breast cancer. This study will be presented in Chapter \ref{chap:wavelet}.


In another project, we focused on a particular type of biological network namely signaling pathway network. Based on a modeling framework of cell signaling, gene expression profiles can be translated into personalized profiles of signaling pathway activities by integrating signaling pathway network. When gene-level profiles are replaced by these derived pathway-level profiles as input to many off-the-shelf computational tools, a simple scheme emerges where gene-level analysis is easily promoted to pathway-level analysis. Analysis made at the level of functional pathways consequently leads to better understanding in terms of biological relevance. Notably, when combined with feature selection oriented algorithms, the proposed scheme enables direct identification of pathway-level mechanistic signatures as an alternative to conventional gene-based signatures, which provides more informative insights into the cellular functions involved in cancer survival. This study will be presented in Chapter \ref{chap:hipathia}.


\paragraph{Other Contribution.}

During the course of my doctoral studies, I have undertaken some other projects as well. In 2013, Elsa Bernard, Erwan Scornet, V\'{e}ronique Stoven, Thomas Walter, Jean-Philippe Vert and I from our laboratory participated as a team in an international bioinformatics competition called \textit{DREAM 8 NIEHS--NCATS--UNC Toxicogenetics Challenge}. In the competition, participants were asked to predict the response of human cell lines exposed to various toxic chemical compounds, based on the chemicals' molecular characterization and the cell lines' transcriptome. Finally our team won second place with a kernel bilinear regression model. Oral presentation was accepted to \textit{NIPS Workshop on Machine Learning in Computational Biology (MLCB)} and later invited to \textit{RECOMB Conference on Regulatory and Systems Genomics}. This work has been published as part of the crowd-source collaboration as a result of the competition \cite{Eduati2015Prediction} but excluded from the present thesis.


During the spring of 2015 when I was interning at Roche Diagnostics, Penzberg, Germany, I worked on failure state prediction for automated analyzers for analyzing biological samples, and a European patent application regarding the methodology was filed by Roche in December 2016 and is currently pending approval \cite{Jiao2016Failure}. Due to corporate confidentiality policies, this study will not be included in the present thesis.
